---
title: "Exercise_06"
date: "13th October 2023"
status: "Completed"
---

```{r}
library(tidyverse)
library(sf)
library(spData)
```

### BMC Regional GIS Data on ACS 2013 - 2017 Quick Facts for Census Tracts

Data Source: [BMC Regional GIS](%22https://gisdata.baltometro.org/datasets/BMC::acs-2013-2017-quick-facts-for-census-tracts/about%22)

American Community Survey(ACS) Quick facts 2013 - 2017 for Census Tracts

This data set was aggregated by Baltimore Metropolitan Council for use in demographic profiles and applications.

It involves features across population statics, employment, rent, mortgage, transportation, education, and location.

### Reading the GeoJSON file into RStudio

```{r}
census_facts <- read_sf("https://gis.baltometro.org/arcgis/rest/services/Census/ACS2013_2017_QuickFacts/MapServer/1/query?outFields=*&where=1%3D1&f=geojson")
census_facts
```

### Why is this data untidy?

-   This data set could be called "untidy" for the columns that are repeated. For eg.: There is one column named "PERCENT_INCOME_BELOW_POV_LEVEL" and another "PERCENT_INCOME_ABOVE_POV_LEVEL". Also,

PERCENT_INCOME_ABOVE_POV_LEVEL = 100 - PERCENT_INCOME_BELOW_POV_LEVEL

-   Also, there are separate columns for number of people belonging to a race and other for percentage of people in those races.

```{r}
attributes(census_facts)
#View(census_facts)
```

### Plotting using ggplot

```{r}
library(ggplot2)
```

Let's plot a map that shows the demographic difference in distribution of people of different ethnicity

We will observe that we cannot demonstrate a visual difference using this visualization as all the regions have almost the same percent of ethnic people

```{r}
ggplot() +
  #geom_sf(aes(fill = "Percent_White"), census_facts) +
  geom_sf(aes(fill = "Percent_Black"), census_facts) #+
  #geom_sf(aes(fill = "Percent_Asian"), census_facts) 
  #geom_sf(aes(fill = "black"Percent_Hispanic), census_facts)
```

Using the pivot_longer() function, the percentage of ethnicity are collapsed into a single column.

The Advantages of lengthening this sf object -

-   We can procure a different row for ethnicity on further cleaning.

-   Using a simple group_by() function,we can use the new data frame to understand the variation in ethnicity.

```{r}
census_facts |>
  pivot_longer(
    cols = c("Percent_White", "Percent_Black", "Percent_Asian", "Percent_Hispanic"),
    names_to = "Ethnicity",
    values_to = "Percent_of_ethnicity"
  )
```

```{r}
census_facts |>
    pivot_longer(
    cols = c("Percent_White", "Percent_Black",  "Percent_Hispanic", "Percent_Asian"),
    names_to = "Ethnicity",
    values_to = "Percent_of_ethnicity"
  ) |>
  group_by(OBJECTID, Ethnicity) |>
  summarize(
    max_pop = max(Percent_of_ethnicity)
  )|>
  ggplot()+
  geom_sf(aes(fill =max_pop, color = Ethnicity))
```

In the above code, the color aesthetic is overridden by the last Ethnicity. However, the max_pop, i.e., the population percent of the ethnicity with the largest number of people in each region, is highlighted on a gradient scale. This map can be useful to interpret only if we know which ethnicity is the majority in each region.

#### The census_facts data frame is a pretty tidy data set even though it has a lot of information, i.e., 53 features. The tidiness of the data makes the analysis and interpretation easier.
